---
title: "Module 5: Advanced Visualization with ggplot"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "", prompt = TRUE, collapse = TRUE, tidy=FALSE, fig.align = "center")
```


### Recap


In this module, we dive into more advanced data visualization with the package `ggplot2`. The `gg` stands for "grammar of graphics", a framework for data visualization. This framework separates the process of visualization into different components: data, aesthetic mappings, and geometric objects. These components are then added together (or layered) to produce the final graph.

We load the `tidyverse` package and the diving dataset from Lecture 1. 
```{r, message = F}
library(tidyverse)
diving <- read_csv("data/diving.csv")
```

1. Data

The first step is to tell `ggplot` what data you will be graphing. This is done using the call to `ggplot` below. We save our graph as `hist`, to which we will then `+` layers. 

```{r}
hist <- ggplot(data = diving)
```

*Note*: if you change your dataset, you need to do this step again.

2. Aesthetics (`aes`)

Aesthetics map the data to the properties of the plot. Examples include:

* `x`: the variable that will be on the x-axis
* `y`: the variable that will be on the y-axis
* `color`: the variable that categorizes data by color
* `shape`: the variable that categorizes data by shape

You can define the `aes` in the `ggplot` call, which will then be used for all later layers, or you can define the `aes` in the `geom` (see below), which will only apply to that `geom`.

<!-- We should show how to change color, shape, and size of points in scatterplots based on a factor. One way to do so would be with the nba_boxscore.csv file. We can plot the shooting percentages based on the player position. Could also compute average stats for each team.
Plot shooting percentages where we colour based on position and change the point size based on number of attempts 
-->
3. Geometric Objects (`geoms`)

Geometric objects, or `geoms`, determine the type of plot that will be created. Examples include:

* `geom_point()`: creates a scatterplot
* `geom_histogram()`: creates a histogram
* `geom_line()`: creates a line
* `geom_boxplot()`: creates a boxplot

Putting it all together:

```{r}
hist <- hist + geom_histogram(aes(x = JScore), binwidth = 0.25)
```

We can then print the graph by seeing the output of `g`:
```{r}
hist
```

We can also add a layer to change the axis labels:
```{r}
hist <- hist + labs(x = "Judge Score")
hist
```

### Facets

What if we want to separate the judges' scores by country? We can use *facets*. Facets allow you to separate graphs by category. We do not need to redo our above code for the histogram - we only need to add a *facet* layer to our graph `hist`. The first argument of `facet_wrap` is the column of our dataset that contains the category information. Note that we need a tilde (`~`) in front of `JCountry`: later in the course we will discuss why this is needed. The second argument of `facet_wrap` specifies the number of rows for which to display the graphs.

```{r, fig.width = 10, fig.height = 5}
hist <- hist + facet_wrap(~ JCountry, nrow = 3)
hist
```

Let's plot the median diving score over the histograms by adding a `geom_vline`. This adds a vertical line (i.e. vline) to your plot.

```{r, fig.width = 10, fig.height = 5}
median_score <- median(diving[["JScore"]])
hist <- hist + geom_vline(xintercept = median_score, color = "red")
hist
```


*Note*: `color` is not in `aes` in the above code because it is a single value. In the next example, each category has a different color, so color is in `aes`.

**Exercise**

Create a histogram of the diving scores with your own choice of category for `facet_wrap` (i.e. not by country).

### Boxplots

Boxplots are also an important visualization tool. We now create boxplots of the judges' scores, separated by diving round. Here we use the `aes` property `fill = Round` to color (or "fill") the boxplots by country.

```{r}
box <- ggplot(data = diving)
box <- box + geom_boxplot(aes(x = Round, y = JScore, fill = Round))
box <- box + labs(title = "Diving scores by round", x = "", y = "Judge Score", fill = "Round")
box
```

We can also remove the unnecessary x-axis ticks and labels as the legend on the right is sufficient. We do so using the `theme` layer:

```{r}
box <- box + theme(axis.title.x = element_blank(), axis.text.x = element_blank(), axis.ticks.x = element_blank())
box
```

### Barplots

We can also create barplots using `geom_bar`:
```{r, fig.width = 10, fig.height = 5}
bar <- ggplot(data = diving)
bar <- bar + geom_bar(aes(x = JCountry, fill = JCountry))
bar <- bar + labs(x = "Judge Country", fill = "Judge Country")
bar
```


### Scatterplots

```{r, include = F}
coeff = lm(JScore ~ Rank, data = diving)$coefficients
round(coeff, 1)
# (Intercept)      JScore 
#       8.1        -0.1 
```

Now let's turn back to scatterplots, which were introduced in [Module 2](module2.html). We plot judges' score versus rank of the diver. As we expect, the higher the divers' rank, the higher their score.

```{r}
scatter <- ggplot(data = diving)
scatter <- scatter + geom_point(aes(x = Rank, y = JScore, color = Country))
scatter
```

### Lines

We can also add an `abline` to our scatterplot - that is, a line where we specify the y-intercept (`intercept`) and the slope (`slope`):

```{r}
scatter <- scatter + geom_abline(intercept = 8.1, slope = -0.1, color = "red")
scatter
```

*Note*: `geom_abline` is different from `geom_line`: `geom_line` "connects the dots" between your data and so doesn't have to be a straight line, whereas `geom_abline` draws a straight line with the specified slope and y-intercept.

Other `geom` for lines are:
- `geom_vline`: to add a vertical line to a plot
- `geom_hline`: to add a horizontal line to a plot

### Stats

We can also specify a layer using `stat_`, which stands for statistical transformation. This is useful if we want to plot a summary statistic of our data, such as a mean or median. By using a `stat_` layer, we do not have to compute this summary statistic beforehand - `ggplot` will compute the summary statistic for us and then plot the result. 

For example, suppose we want to plot the means of each judge's score and provide error bars of one standard deviation on either side of the mean. We could use `summarize` and `group_by` to find the mean and standard deviations for each judge, *or* we could just use a `stat_` layer!

The layer `stat_summary()` computes and then plots a user-specified summary statistic. We choose the option `mean_se` to calculate the means and standard deviations of the scores of each judge.

As always, we set up the plot by calling `ggplot`, specifiying `data = diving` and then providing the `aes`. In this case, we want the judge on the `x`-axis and their scores on the `y`-axis. We then add out `stat_summary` layer.

```{r}
judges <- ggplot(data = diving, aes(x = Judge, y = JScore))
judges <- judges + stat_summary(fun.data = mean_se)
judges <- judges + labs(y = "Judge Score")
judges
```

We can see that the judges' names are bunched together... ideally we would want their names to be horizontal to make it easier to read. Luckily, `ggplot` has an easy fix for this!

```{r}
judges <- judges + coord_flip()
judges
```


### Scales

Scales allow you to adjust the aesthetics or visual aspects of a plot.  We return to the scatter plot of the judges' scores vs rank of the divers. This time, we want to color the points by the difficulty of the dive.

We use the layer:

- `scale_color_distiller`

Above, the second word, `color`, is the `aes` we want to change.  We can replace it with `x`, `y` or `fill`, depending on the `aes` we want to change. 

The third word is `distiller`, which we use because our `color` variable, `Difficulty`, is continuous. If it were discrete, we would write `brewer` instead.

```{r}
scatter <- ggplot(data = diving)
scatter <- scatter + geom_point(aes(x = Rank, y = JScore, color = Difficulty))
scatter <- scatter + scale_color_distiller(palette = "OrRd", direction = 1)
scatter
```

Interestingly, it seems that some of the highest ranked divers perform most of the less difficult dives, but perform these easy dives very well. 

To further investigate, we plot a 2D histogram of Rank vs Difficulty.

```{r}
hist <- ggplot(data = diving)
hist <- hist + geom_bin2d(aes(x = Rank, y = Difficulty))
hist <- hist + scale_fill_distiller(palette = "Spectral")
hist
```

Note that we use `fill` instead of `color` in `scale_fill_distiller` to control the fill of the histogram bins.

From the 2D histogram, we can see that the higher ranked divers attempt both more difficult and less difficult dives, unlike the lower ranked divers who only attempt more difficult dives. 

### ColorBrewer

The palettes used in this module, including "OrRd" and "Spectral", come from [ColorBrewer](http://colorbrewer2.org/). You can take a look at the website and use some of these color palettes in your plots!



