---
title: "Lecture 3: Visualizations with categorical data and grouped operations"
output: html_document
---

```{r lecture3-setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "", prompt = TRUE, collapse = TRUE, tidy=TRUE, fig.align = "center")
library(tidyverse)
```

We're going to pick up right where we left off in [Lecture 2](tc_lecture2.html) by
loading the `nba_shooting` dataset we created using the `filter` and `mutate` functions.

```{r load}
load("data/clean_nba_shooting.RData")
```

We're interested in the rate and efficiency of three-point attempts over time. To start,
let's make simple histograms for `three_point_fg_rate = TPA / FGA` and `TPP` following
what we learned in [Lecture 2](tc_lecture2.html).

```{r tp-rate-hist}
nba_shooting %>%
  ggplot(aes(x = three_point_fg_rate)) +
  geom_histogram(binwidth = 0.05) +
  # label!
  labs(x = "Three-point shot rate = TPA / FGA",
       y = "Count",
       title = "Distribution of three-pointer shot rate")
```

```{r tpp-hist}
nba_shooting %>%
  ggplot(aes(x = TPP)) +
  geom_histogram(bins = 30) +
  # label!
  labs(x = "Three-point shot proportion = TPM / TPA",
       y = "Count",
       title = "Distribution of three-point shot proportion")
```

Next let's make a scatterplot showing the relationship between the rate at which
players attempt three point shots and the proportion at which they make them:

```{r tp-scatter}
nba_shooting %>%
  ggplot(aes(x = three_point_fg_rate, y = TPP)) +
  geom_point(alpha = 0.25) +
  labs(x = "Three-point shot rate = TPA / FGA",
       y = "Three-point shot proportion = TPM / TPA",
       title = "Relationship between three point shot rate and efficiency")
```

Unsurprisingly, for the most part we see a random scattering of points. But we can see that
players who attempt a minimal proportion of three-point shots for their field goal attempts
also have typically low values for their three point shot efficiency. You can expect players 
that can make three-pointers more often to then attempt more and more three-pointers - but,
as the plot shows, they will reach a point where their advantage diminishes.

## Plotting categorical data

The style of shooting in the NBA has been evolving over the past decade, with teams
attempting more and more three-point shots than ever before. We can look into this trend
ourselves using the `nba_shooting` dataset we constructed. Something such 
as the season/year could be considered like a continuous variable. For instance, 
we could create a scatterplot with the `SEASON` variable on the x-axis and each
of the variables of interest on the y-axis, such as:

```{r season-scatter-tp-rate}
nba_shooting %>%
  ggplot(aes(x = SEASON, y = three_point_fg_rate)) +
  geom_point(alpha = 0.25) +
  labs(x = "Season",
       y = "Three-point shot rate = TPA / FGA",
       title = "Distribution of three point shot rate each year")
```

It is hard to see any real differences between the years this way. Instead, we
can treat the season as a __categorical variable__ to then more appropriately
compare the distributions between each year. To do so, we are going to create a
new variable in our dataset called `season_factor` which is simply a new version 
of the `SEASON` variable but is converted to a special type of data in `R` known
as a "factor". We won't cover the details, but "factors" are essentially categorical
variables with an order to them. By default, when we create `season_factor` using 
the `as.factor()` function, the natural chronological order is used.

```{r season-factor}
nba_shooting <- nba_shooting %>%
  mutate(season_factor = as.factor(SEASON))
```

Now with this variable, prior to comparing distributions between
the different seasons, we will create a __barchart__ to compare the number of players
we have in each season that met our criteria from [Lecture 2](tc_lecture2.html).

This can be done using `geom_bar`, and all we have to specify is the categorical
variable to be displayed along the x-axis - `ggplot` will count the number of each
player and display it for us just like `geom_histogram`.

```{r season-bars}
nba_shooting %>%
  ggplot(aes(x = season_factor)) +
  geom_bar() +
  labs(x = "Season", y = "Count", 
       title = "Number of players in each season meeting TPA and FGA cutoffs",
       subtitle = "Excluding lockout seasons 1999 and 2012")
```

What shouldn't come as a surprise is that we're seeing more and more player stats
in more recent years. This makes sense since more players are shooting three-pointers
more frequently now. 

## Comparing distributions

Now that we have a categorical version of the season to use for plotting, we can
proceed to compare the distributions of the three-point stats in a variety of ways.

Naturally you can make __multiple__ histograms to compare the distributions of several 
different categories with simple extensions of code you are already familiar with.
This can be done in several different ways.

### Facets

We can use *facets* to create histograms for each different season. Facets allow you to separate graphs by category. 
We do not need to do anything different from our above code for creating histograms except we only need to add a *facet* layer to our graph. The first argument of `facet_wrap` is the column of our dataset that contains the category information.Note that we need a tilde (`~`) in front of `season_factor`. The second argument of `facet_wrap` specifies the number of rows for which to display the graphs.

```{r facet-hist}
nba_shooting %>%
  ggplot(aes(x = three_point_fg_rate)) +
  geom_histogram(binwidth = 0.05) +
  # label!
  labs(x = "Three-point shot rate = TPA / FGA",
       y = "Count",
       title = "Distribution of three-pointer shot rate by season") +
  facet_wrap(~season_factor, nrow = 3)
```

Although we can now see a histogram of three-point shot rate for each separate season,
it's a little hard to make comparisons across all the years. 

### Colors

We can also proceed to create multiple histograms using colors instead of facets,
by just adding the `fill = season_factor` line to the `aes` function. **NOTE**: it's
important to remember the difference between `fill` and `color`. In this case, `fill`
actually changes the literal _fill_ of the bars while `color` changes the line. This
differs from `geom_point` and is important to keep in mind when making plots.

```{r hist-color-stacked}
nba_shooting %>%
  ggplot(aes(x = three_point_fg_rate, fill = season_factor)) +
  geom_histogram(binwidth = 0.05) +
  # label!
  labs(x = "Three-point shot rate = TPA / FGA",
       y = "Count",
       fill = "Season",
       title = "Distribution of three-pointer shot rate by season") 
```

Well that's a weird looking plot! This is because by default the multiple 
histograms will be stacked on top of one another. We can instead change the 
`position` of the `geom_histogram` call to be equal to "identity" rather than
the default which is "stacked". When doing this we'll have several histograms
overlaid on top of each other, so it's important to change the `alpha` to be
a lower value.

```{r hist-color-dodge}
nba_shooting %>%
  ggplot(aes(x = three_point_fg_rate, fill = season_factor)) +
  geom_histogram(binwidth = 0.05, position = "identity",
                 alpha = 0.1) +
  # label!
  labs(x = "Three-point shot rate = TPA / FGA",
       y = "Count",
       fill = "Season",
       title = "Distribution of three-pointer shot rate by season") 
```


Overall this is still a rather difficult figure to digest. Instead, we should 
make figures that take advantage of the natural ordering implied by the season
along an axis while displying various comparisons of the distributions.

## Boxplots

By far the simplest way to compare distributions across many levels of a categorical
variable is with boxplots. Side-by-side boxplots can be generated easily using
`geom_boxplot`. Now we specify the categorical variable we want __mapped__ to the
x-axis (`season_factor`) and then the variable we want mapped to the y-axis.

```{r boxplot}
nba_shooting %>%
  ggplot(aes(x = season_factor, y = three_point_fg_rate)) +
  geom_boxplot() +
  # label!
  labs(x = "Season",
       y = "Three-point shot rate = TPA / FGA",
       title = "Comparison of three-pointer shot rate for each season with boxplots") 
```

Remember, boxplots are displaying summary statistics (median, min, max, and percentiles)
which allows us to easily see an increase in the median three-point shot rate over
time. **But boxplots are extremely limited visualizations!** All they provide are
summary statistics - think of all the information we observed in histograms that
are lost in boxplots! There is no way of knowing if we're looking at distributions
with multiple modes!

## Violin plots

A much more informative plot than only a boxplot, is a __violin plot__. Violin 
plots display the density curves giving us the general shape of a distribution. 
And of course, there's a `geom` for that as well called... `geom_violin`!

```{r violin}
nba_shooting %>%
  ggplot(aes(x = season_factor, y = three_point_fg_rate)) +
  geom_violin() +
  # label!
  labs(x = "Season",
       y = "Three-point shot rate = TPA / FGA",
       title = "Comparison of three-pointer shot rate for each season with violin plots") 
```

This gives us a sense of the distributions in each season, and any sort of subtle
differences over time. What's missing from this plot is the key points provided
by boxplots to show the increasing rate of three point attempts. Because we are
using `ggplot`, we can easily add the boxplot layer on top of the violin plots
to get the best of both visualizations! We just specify a small `width` for our
boxplots so that they don't take up too much space and fit inside the violin plots.

```{r violin-box}
nba_shooting %>%
  ggplot(aes(x = season_factor, y = three_point_fg_rate)) +
  geom_violin() + 
  geom_boxplot(width = 0.2) +
  # label!
  labs(x = "Season",
       y = "Three-point shot rate = TPA / FGA",
       title = "Comparison of three-pointer shot rate for each season with violin and boxplots") 
```

